---
title: 'Mini Project 1'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author: "Instructor: Ali Abuzaid"
format: 
 pdf:
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

::: callout
**STUDENT NAME**

-   Xinyao Song (6194773)
-   Jay Chang (5808498)
-   Nic Chan (A052038)
:::

::: callout-tip
# Instructions

-   This mini project aims to familiarize you with real-life data sourced from various resources.

-   The mini project includes narrative questions. While these questions are primarily based on lecture material and prerequisites, they may also require independent thinking and investigation.

-   Collaborate in groups of **2-3** students from the **same section**; individual submissions **will not be accepted**.

-   Please use the provided `MP 1.qmd` file to type your solutions and submit the completed assignment as a PDF file. You can utilize `RStudio`for this purpose. For guidance, refer to the [Tutorial: Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)).

-   Submit your answers via **Gradescope**.

-   Ensure that all `R` code, mathematical formulas, and workings are presented clearly and appropriately.

-   All figures should be numbered, and axes must be labeled.
:::

::: callout-caution
# Due Date

**Due Date:** Sunday, May 4, 2025, 11:59 PM
:::

\newpage

::: callout-tip
# Survey (Use the attached 'Survey.xlsx' file)

The data was collected during the first lecture from 85 students in the PSTAT 100 class in Spring 2025. The survey has 6 main sections namely Personal Information, Physical Measurements, Health Habits, Diet and Nutrition, Mental Health and Stress, and Academic Life.

These sections are interconnected - for example, physical measurements may relate to health habits and diet, while academic life may impact mental health and sleep patterns.The goal is to explore the relevant variables that may affect students' performance.
:::

::: callout-important
# **Question 1**

Based on your understanding of the collected data, propose three questions or hypotheses that can be explored or tested using the data.
:::

`r bfcolor("ANSWERS TO QUESTION 1:", "red")`\

**Hypothesis 1:** Whether the amount of sleep a student gets is associated with their GPA.\
$H_0:$ The amount of sleep a student gets does not affects with their GPA ($\mu_{\text{<4 hours}} = \mu_{\text{4-6 hours}} = \mu_{\text{6–8 hours}} = \mu_{\text{>8 hours}}$).\
$H_A:$ The amount of sleep a student gets does not affects with their GPA (at least one sleep group has a different mean GPA).

**Hypothesis 2:** Whether the amount of physical exercise a student does affects their sleep duration.\
$H_0:$ There is no association between the amount of exercise and the amount of sleep ($\text{Exercise} \perp \text{Sleep}$).\
$H_A:$ There is an association between the amount of exercise and the amount of sleep ($\text{Exercise} \not\perp \text{Sleep}$).

**Hypothesis 3:** Whether a student's stress level influences their likelihood of consuming alcohol.\
$H_0:$ There is no association between stress level and alcohol consumption ($\text{Stress} \perp \text{Alcohol}$).\
$H_A:$ There is an association between stress level and alcohol consumption ($\text{Stress} \not\perp \text{Alcohol}$).

\newpage

::: callout-important
# **Question 2**

Load the data into `R`, obtain basic descriptive statistics for the numerical variables, and provide your interpretation of the results.
:::

`r bfcolor("ANSWERS TO QUESTION 2:", "red")`\

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
survey <- read_excel("Survey.xlsx")
numeric_var <- survey %>% select(where(is.numeric))
summary(numeric_var)
```

Among all the variables in the Survey dataset, there are a total of four numerical variables, which include weight, height, stress, and GPA.

-   From the summary statistics above, we can see that the weight is likely measured in pounds, with the minimum weight being 106 pounds and the maximum being 390. We can see that there is a significant difference between the third quartile and the maximum weight value, this may suggest that the data is right-skewed or that there are outliers on the higher end.
-   On the other hand, the height variable seems a little inconsistent in formatting given the minimum value being 5 and the maximum value being 511. It's possible that most data were recorded in feet-inches, and that entries like 511 were meant to represent "5 feet 11 inches" but were entered without a decimal. There is also a huge spread between the third quartile and the maximum value, which may suggest that there are multiple entry errors that need to be fixed.
-   While the stress levels are divided into 10 levels, we can see that they are rather even throughout each quartile. One thing to note is that while there is a maximum stress level of 10, the minimum stress inputted is 2, not 1.
-   Finally, for GPA, we can see that nearly 75% of the data lies above a 3.3 GPA. There seems to be a huge gap between the minimum value and the 1st quartile. Since this is a survey, we might need to take a closer look into voluntary bias, as students with lower GPA may not be willing to share their data.

\newpage

::: callout-important
# **Question 3**

a- Identify any missing values in the `Survey` dataset.

b- Assess the missing data mechanism in the dataset.

c- How would you handle the identified missing values? Save the treated dataset as `SurveyMissing`?

d- Can you address any potential bias for this missing values?
:::

`r bfcolor("ANSWERS TO QUESTION 3:", "red")`\

a.  

```{r}
colSums(is.na(survey))

```

From the summary above, we see that there are 2 missing values in Alcohol, 3 missing values in GPA, and 5 missing values in Health.

b.  The missing data for GPA is likely due to MNAR, as there is a huge gap between the minimum value and the first quartile, and students with lower GPAs are less likely to report it due to pressure or personal reasons. On the other hand, the missing values in the Health variable are most likely due to MAR, as the survey question is "Do you feel your academic workload affects your health?". This question may have some students finding difficult to answer, which may result them in skipping the question. Nevertheless, this missing value in Health may potentially be explained by the participants' GPA or stress levels. Finally, the missing values in Alcohol may be explained by either MNAR or MAR. It is MNAR as some students may not want to share their alcohol consumption due to their age in case of underage drinking or concerns with their social environment. Alternatively, it can also be MAR if the variable can be related to stress or other observed factors.

c.  Since GPA is additive, bounded, and MNAR with a very small proportion of missing values, we will use arithmetic mean imputation to handle the identified missing values as GPA. In terms of the two categorical variables, Alcohol and Health, which are both and potentially MAR, mean imputation or dropping the data would potentially induce bias in the dataset. Some of the better ways to handle it are through inverse probability weighting or modeling the variables with missing observations as functions of other variables. However, since these methods are not covered explicitly, we will mark the missing variables with "Prefer not to Answer" for later analysis that suits the type of the variable.

```{r}
SurveyMissing <- survey %>%
  mutate(
    GPA = ifelse(is.na(GPA), mean(GPA, na.rm = TRUE), GPA),
    Alcohol = ifelse(is.na(Alcohol), "Prefer not to answer", Alcohol),
    Health = ifelse(is.na(Health), "Prefer not to answer", Health)
  )
```

d.  Since we used mean imputation for GPA by assuming GPA is most likely MCAR, it may still distort the distribution and introduce bias into the dataset. For Alcohol and Health that are categorical, we did not explicitly handle the missing values, as dropping those rows would introduce more bias. Instead, we marked the missing values with "Prefer not to Answer," which preserves the structure of the data and allows us to retain those rows without making assumptions about their true values, enabling better analysis in the future.

\newpage

::: callout-important
# **Question 4**

a- Detect and handle outliers in the `Weight` variable of the dataset.

i- Justify your outlier detection method (e.g., IQR, Z-score, or visual inspection) and explain your chosen handling strategy (e.g., winsorization, capping, or removal).

ii -Discuss the potential impact of these outliers on the analysis if left unaddressed.
:::

`r bfcolor("ANSWERS TO QUESTION 4:", "red")`\

Before answering Question 4, we will address the messy/inconsistent 'Height' values. Upon looking at it, we noticed that many values are incorrect, inconsistent, and very unclear, with some missing decimals and others representing unrealistic values when comparing them to the person's weight and gender, such as a male who weighs 116 pounds, yet is listed as '66' (which we need to determine means 6 foot 6 inches or 66 inches).

To clean this data, we have determined several guidelines to correct and convert all values to inches:

1.  If the height listed has a decimal, then assume its in standard feet/inches format, and convert to inches.
2.  If the height listed is a single number (i.e. 6), then assume its in feet and convert to inches.
3.  If the height listed is greater than 500, then assume it's in feet/inches format and missing a decimal, and convert to inches (i.e. 511 should be 5'11").
4.  If the height listed is greater than 100 but less than 500, then assume it's in centimeters and convert to inches.
5.  If the height listed is 61 or less, then assume that it's missing a decimal and convert to inches (i.e. 56 means 5'6").
6.  If the height listed is greater than 61, then assume that it's already in inches, so no conversion needed.

```{r}
convert_height <- function(h_raw) {
  suppressWarnings({h_raw <- as.numeric(h_raw)
    
    if (is.na(h_raw)) {
      return(NA)
    } else if (h_raw %% 1 != 0) {
      # Rule 1: Contains a decimal, feet.inches format
      ft <- floor(h_raw)
      inch <- round((h_raw - ft) * 10)  # e.g. 5.7 → 5'7"
      return(round(ft * 12 + inch))
    } else if (h_raw <= 8) {
      # Rule 2: single-digit number, feet
      return(round(h_raw * 12))
    } else if (h_raw > 500) {
      # Rule 3: >500, missing decimal, feet.inches
      ft <- floor(h_raw / 100)
      inch <- h_raw %% 100
      return(round(ft * 12 + inch))
    } else if (h_raw > 100 && h_raw <= 500) {
      # Rule 4: 100–500, cm to inches
      return(round(h_raw * 0.393701))
    } else if (h_raw <= 61 && h_raw > 8) {
      # Rule 5: ≤61, missing decimal, feet.inches
      ft <- floor(h_raw / 10)
      inch <- h_raw %% 10
      return(round(ft * 12 + inch))
    } else {
      # Rule 6: >61, already inches
      return(round(h_raw))
    }
  })
}

# Applying function to DataFrame
survey <- survey %>%
  mutate(Height_cleaned = sapply(Heigh, convert_height))
  
# Displaying cleaned height data vs original heights
survey %>%
  select(Heigh, Height_cleaned) %>%
  head(20) 
```

Now, we can see that all our height values under 'Height_cleaned' are correctly converted to inches, are realistic values based on the rest of the data, and uniform.

Now, we can move onto answering Question 4.

a.  When it comes to addressing any outliers in the 'Weight' variable of the dataset, we can detect it using a combination of calculating the IQR and visual inspection and handle it by winsorization/capping.

```{r}
# Initial Summary
summary(survey$Weight)

# Calculating IQR
Q1 <- quantile(survey$Weight, 0.25, na.rm = TRUE)
Q3 <- quantile(survey$Weight, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Find outliers based on IQR/bounds
outliers <- survey %>% filter(Weight < lower_bound | Weight > upper_bound)
print(outliers$Weight)

# Winsorization: capping extreme values
survey <- survey %>%
  mutate(Weight_Capped = ifelse(Weight == 390, upper_bound, Weight))
```
Although we found 250 as one of the outliers based on IQR, upon further visual inspection we see that the value of 250 could be a realistic weight value, especially when comparing it to the height and gender of this particular person (6'4" male). As a result, we have determined that the value of 390 is the only outlier in our dataset, and have handled it accordingly.

i.  Through a simple visual inspection, we can see that there aren't that many egregious weight values at all, so this is a very fast and simple way of detecting potential outliers. At the same time, visual inspection allows for us to analyze the data in a deeper, more realistic/personal way, potentially noticing different aspects to our data that could explain inconsistencies or outliers. Additionally, by calculating the IQR (which doesn't get distorted by extreme values), we are able to find the different quartiles and set a lower and upper bound, allowing for easy detection of outliers. Also, we can use the IQR to make visual representations like boxplots that help see outliers. For our handling strategy, we chose to use winsorization/capping because it allows us to keep all data points and maintaining the original sample size. Also, it reduces skewness and the influence of extreme values on sample statistics such as the mean and regressions. 

ii. Leaving these outliers can have several adverse effects on our analysis. It can skew summary statistics, where a single high outlier can shoot the mean upward, making it seem like the "typical" or average weight is higher than it actually is. It can also inflate the variance and standard deviation, affecting things like confidence intervals and hypthosis testing. Lastly, it can impact different correlations, such as between Weight and Stress.

\newpage

::: callout-important
# **Question 5**

Investigate the relationship between sleep duration (`Sleep`) and perceived stress levels (`Stress`). Address the following:

a- Calculate the average stress level for each sleep duration category and comment.

b- Do certain sleep ranges (e.g., "4–6 hours") correlate with higher/lower stress?

c- Use a boxplot to compare stress distributions across sleep categories and comment!
:::

`r bfcolor("ANSWERS TO QUESTION 5:", "red")`\

a. Our first step in investigating the relationship between sleep duration and perceived stress levels is calculating the average stress level for each sleep duration category.

```{r}
# Ensuring stress is numeric
survey <- survey %>%
  mutate(Stress = as.numeric(Stress))

# Grouping by each Sleep category and calculating avg stress
sleep_stress_avg <- survey %>%
  filter(!is.na(Sleep), !is.na(Stress)) %>%
  group_by(Sleep) %>%
  summarise(
    avg_stress = round(mean(Stress), 2),
    count = n()
  ) %>%
  arrange(Sleep)

print(sleep_stress_avg)
```
At face value, we can see that the average stress levels vary from category to category, but not too much, and doens't dip under 5. We can also see that there seems to be a correlation between amount of sleep and stress levels, but we will address this in the next part. 
 
b. As we can see from the summary of average stress level for each 'Sleep' category, the results make sense, and we can see that there is a clear correlation between certain sleep ranges and higher/lower stress levels. People sleeping for less than 6 hours have higher overall average stress levels, with those sleeping less than 4 hours having the highest stress levels. In contrast, those sleeping a sufficient amount (more than 6 hours) have lower stress levels, with those sleeping the most (more than 8 hours) having the lowest average stress levels, which makes sense.

c. Now, we can make a boxplot to have a visual representation to compare stress and sleep categories.

```{r}
# Making the boxplot
ggplot(survey, aes(x = Sleep, y = Stress, fill = Sleep)) +
  geom_boxplot() +
  labs(
    title = "Stress Distributions Across Sleep Categories",
    x = "Sleep Duration",
    y = "Stress Level"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
Upon looking at these boxplots, we have a solid visual representation of the stress distributions. Justifying our explanation in the previous part, we can see that the average stress levels are lower for the higher sleep categories (more than 6 hours), and are higher for the sleep categories where they sleep less than 6 hours. Additionally, we can see outliers for the '4-6 hours' sleep category, showing us that some who sleep this amount can have low stress levels, but others can have very high, further justifying that fewer hours of sleep correlates to higher stress levels and vice versa.

\newpage

::: callout-important
# **Question 6**

Create a visualization to show the distribution of Weight, comment!
:::

`r bfcolor("ANSWERS TO QUESTION 6:", "red")`\

```{r}
ggplot(data = survey, aes(x = Weight_Capped)) +
  geom_histogram(binwidth = 25,fill = "darkblue", color = "white") + 
  xlab("Weight (lbs)") + ylab("Number of Students") +
  ggtitle("Weight Distribution of Students in PSTAT100") + 
  theme_minimal() +  scale_x_continuous(breaks = seq(100, 300, by = 25))
#Create histogram using the ggplot2 package to show the distribution of weight
#Use Weight_Capped variable to remove outlier
#Binwidth of 25 and x axis scaled by 25 for better interpretability
```
The histogram is made using the Weight_Capped variable which removes the one outlier of 390 pounds. By removing the outlier, the standard deviation and variance of the data is reduced, providing a more accurate representation of the student weight. The histogram reveals a slightly right skewed distribution centered around 150 pounds, which suggests there are some heavier individuals but not enough to significantly alter the data. This data reflects the the average weight of adults ranging from 125 to 175 pounds, which is the bulk of the data.
\newpage

::: callout-important
# **Question 7**

a.  Define a new variable BMI = $\left(\frac{\text{Weight (lb)}}{\text{Height (in)}^2} \right) \times 703$, and classify students into BMI categories based on Centers for Disease Control and Prevention (CDC) guidelines:

-   BMI \< 18.5 → `"Underweight"`\
-   18.5 ≤ BMI \< 25 → `"Normal"`\
-   25 ≤ BMI \< 30 → `"Overweight"`\
-   BMI ≥ 30 → `"Obese"`

b- Visualize and describe the distribution of students' BMI categories using a bar chart.

c- Explore the correlations between **BMI**, **GPA**, and **Stress**. Use visualizations and correlation statistics to summarize the relationships.
:::

`r bfcolor("ANSWERS TO QUESTION 7:", "red")`\

```{r}
#a
survey <- survey %>%
  mutate(BMI = round((Weight_Capped / (Height_cleaned^2)) * 703,2))
survey <- survey %>%
  mutate(BMI_Category = ifelse(BMI < 18.5, "Underweight",
                               ifelse(BMI < 25, "Normal",
                               ifelse(BMI < 30, "Overweight", "Obese"))))
#b
ggplot(survey, aes(x = BMI_Category)) +
  geom_bar(fill = "darkgreen") +
  xlab("BMI Category") + 
  ylab("Number of Students") + 
  ggtitle("BMI Category Distribution") +
  theme_minimal()
```
b. From the barchart it can be seen that the majority of students who took the survey are categorized as "Normal" by the BMI. The next highest category is "Overweight" followed by "Obese" and then by "Underweight". The data shows that most students have a BMI in the recommended range, though it is notable that there are significantly more overweight students compared to underweight students. 

```{r}
#c
# Scatter plot of BMI vs GPA
ggplot(survey, aes(x = BMI, y = GPA)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ggtitle("BMI vs GPA") +
  theme_minimal()

# Scatter plot of BMI vs Stress
ggplot(survey, aes(x = BMI, y = Stress)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ggtitle("BMI vs Stress") +
  theme_minimal()

# Scatter plot of GPA vs Stress
ggplot(survey, aes(x = GPA, y = Stress)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ggtitle("GPA vs Stress") +
  theme_minimal()

#Calculating the correlations between the variables
#Correlation between BMI and GPA
cor(survey$BMI, survey$GPA, use = "complete.obs")

#Correlation between BMI and Stress
cor(survey$BMI, survey$Stress, use = "complete.obs")

#Correlation between GPA and Stress
cor(survey$GPA, survey$Stress, use = "complete.obs")
```
c. I created scatter plots to show the relationships between BMI and GPA, BMI and Stress, and GPA and Stress. I then calculated the Pearson's correlation coefficient. The strongest correlation was about -0.45 in the relation between BMI and GPA, which moderately suggests that students with higher BMI may have lower GPA. The correlation value of 0.14 for BMI vs Stress shows a very weak positive correlation meaning there was only a slight tendency for students with higher BMI reporting more stress. Finally, GPA and Stress had a correlation value of -0.21 which suggests a weak negative correlation showing that students with lower GPA may have higher stress levels. Overall, the strongest correlation was between BMI and GPA while the other relations involving stress were relatively weak which indicate that the variables may be loosely connected and that more data should be pulled to find out more.
